Progress Report

### 2023-09-26 Docs/ProgressReport.Rmd
report on laplace approximation procedure

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
theme_set(theme_minimal())

@@

df %>% 
  filter(SEQN %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Z), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
```

## fGFPCA Estimation

```{r}
# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))

### Binning

- Bin every 10 measures

@@

df %>% 
  filter(SEQN %in% rand_id) %>%
  group_by(SEQN, bin) %>%
  summarise(num = sum(Z)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```

### Local GLMMs

Following up on last report, when using glmer, the results from PRILS (nAGQ=0) seems more in line with experiences. As validation, let's fit the marginal model and look at the estimated population-average latent function, and compare it with the FPCA mean function. Specifically, at bin s the model would be $g(E(Y_i(s)))=f_0(s)$, with no subject-specific random effect.

Another validation I came up with is to look at the local intercepts. That is the $\beta_0(s)$ in local mixed models $g(E(Y_i(s)))=\beta_0(s)+b_i(x)$. This should correspond to the FPCA mean functions, as the population average level.

Since the individual random effects are centered around zero, I believe the marginal $f_0(s)$ should look very much like the conditional $\beta_0(x)$ if the model local fit well. The result obviously is in favor of PRILS.n *But what is the reason for AGQ to overshoot downward?*

```{r}
# split by each bin
df <- df %>% rename(id = SEQN, Y=Z)
df_bin_lst <- split(df, f = df$bin)
```

```{r cache=TRUE}
# local marginal GLM
local_marg <- function(df){
    this_glm <- glm(Y ~ 1, data = df, family = binomial)
    eta_hat <- predict(this_glm, type = "link")
    df$eta_hat <- eta_hat
    return(df)
}

df_est_local_marg <- lapply(df_bin_lst, function(x){local_marg(x)})

# local conditional GLMM
## PIRLS vs AGQ
df_est_cond_mean_pirls <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 0)}) 
df_est_cond_mean_agq <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 5)}) 
```

```{r, fig.width=12, fig.height=4}
ggarrange(
df_est_local_marg %>% bind_rows() %>%
  select(bin, eta_hat) %>% distinct(.) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLM (marginal)", y = "latent mean")+
  ylim(-25, 1),

data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLMM (conditional) using PRILS", y = "latent mean")+
  ylim(-25, 1),

data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_agq)) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLMM (conditional) using AGQ", y = "latent mean")+
  ylim(-25, 1),

nrow = 1, align = "h"
)
```

Let's for now trust the results from PIRLS more than AGQ, if we are using glmer.

```{r cache=TRUE}
# fit local GLMM and estimate latent function
# to avoid near-unidentifiability issues, using PIRLS maximization with nAGQ = 0

# fit local models
t1 <- Sys.time()
df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2 <- Sys.time()
t_pirls <- t2-t1
df_est_latent <- bind_rows(df_est_latent)

# keep midpoint values
df_est_latent <- df_est_latent %>%
  select(-sind, -Y) %>% distinct(.)
```

```{r, message=FALSE}
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r, message=FALSE}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
```

#### An alternative: glmmTMB

The default optimizer of glmmTMB is nlminb (quasi-Newton BFGS), whereas glmer uses bobyaqa (nAGQ=0) or Nelder_Mead (nAGQ>0)?

Let's first try to re-do the who local GLMM step using glmmTMB. Looks like no numeric issue occurred. It takes slightly shorter time (about 1 minute) than PIRLS in glmer

```{r, cache=TRUE}
library(glmmTMB)
# re-fit local GLMM and check numeric issues
try_fit_lst<-list()
t1=Sys.time()
for(i in seq_along(df_bin_lst)){
  # this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
  #                      warning=function(w){NA})
  this_glm <- glmmTMB(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial)
  try_fit_lst[[i]] <- this_glm

}
df_est_latent_tmb <- lapply(try_fit_lst, function(x){predict(x, type = "link")})
t2=Sys.time()
t_tmb <- t2-t1
# population mean
mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
```

- Time: `r round(t_tmb, 2)` minutes
- Mean/Intercept function: It looks like glmmTMB estimates lie right between PIRLS and AGQ! It takes slightly shorter time than PIRLS (`r round(t_pirls, 2)` minutes) and mean function shoots more downward, not as bad as AGQ though.

```{r, fig.width=12, fig.height=4}
ggarrange(
df_est_local_marg %>% bind_rows() %>%
  select(bin, eta_hat) %>% distinct(.) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLM (marginal)", y = "latent mean")+
  ylim(-15, 1),

data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLMM (conditional) using PRILS (glmer)", y = "latent mean")+
  ylim(-15, 1),

data.frame(bin=mid, eta_hat = mean_tmb) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLMM (conditional) using glmmTMB", y = "latent mean")+
  ylim(-15, 1),

nrow = 1
)
```

- Individual latent functions

```{r}
for(i in seq_along(df_est_latent_tmb)){
  df_est_latent_tmb[[i]] <- data.frame(df_bin_lst[[i]], eta_hat = df_est_latent_tmb[[i]])
}

df_est_latent_tmb <- bind_rows(df_est_latent_tmb)

# keep midpoint values
df_est_latent_tmb <- df_est_latent_tmb %>%
  select(-sind, -Y) %>% distinct(.)

@@

```{r, message=FALSE, fig.height=8, fig.width=16}
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
            by = c("id", "bin")) %>%
  mutate_at(vars(eta_hat, eta_hat_tmb), function(x){exp(x)/(1+exp(x))}) %>% 
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, col = "glmer"))+
  geom_line(aes(x=sind, y=eta_hat_tmb, col = "glmmTMB"))+
  geom_point(aes(x=sind, y = Y), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r, message=FALSE, fig.height=8, fig.width=16}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
            by = c("id", "bin")) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, col = "glmer"))+
  geom_line(aes(x=bin, y=eta_hat_tmb, col = "glmmTMB"))+
  geom_point(aes(x=bin, y = num), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function")
```

So it looks like glmmTMB still overshoots downwards. Between the two methods, the mid-day part (active part for most people) seems to be very similar. Difference lies mostly at both ends, with consecutive inactive minutes.

```{r, eval=FALSE}
# numeric values of individual latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
            by = c("id", "bin")) %>%
  head(20)
```

Follow up from late time, let's use PIRLS estimator and fix the number of PC functions instead of using all of them. We start with four PC functions.

I will use estimated latent functions from both glmer (nAGQ=0) and glmmTMB, and compare their results.

```{r}
# glmer nAGQ=0
mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)

# lmmTMB
mat_est_unique2 <- matrix(df_est_latent_tmb$eta_hat, 
                          nrow=N, ncol=n_bin, byrow = F) 
fpca_mod2 <- fpca.face(mat_est_unique2, argvals = mid, var=T)

# ncol(fpca_mod$efunctions) # 27 eigenfunctions
K <- 4
```

- Population mean function

Hmm...Compared to conditional intercept, FPCA seems to shoot the population mean downward in PIRLS but upward for glmmTMB. Does that make sense?

```{r}
# population mean
par(mfrow=c(1, 2))
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmer",
     ylim = c(-12, 1))
plot(mid, fpca_mod2$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmmTMB", ylim = c(-12, 1))
```
- PC functions

```{r, fig.height=16, fig.width=8}
# eigenfunctions
par(mfrow=c(4,2))

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmer", ylim = c(-0.21, 0.13))
plot(mid, fpca_mod2$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmmTMB", ylim = c(-0.21, 0.13))

plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2", ylim = c(-0.2, 0.14))
plot(mid, fpca_mod2$efunctions[, 2], type="l", xlab="bin", ylab="PC2",ylim = c(-0.2, 0.14))

plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3", ylim = c(-0.2, 0.2))
plot(mid, fpca_mod2$efunctions[, 3], type="l", xlab="bin", ylab="PC3",ylim = c(-0.2, 0.2))

plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4", ylim = c(-0.14, 0.23))
plot(mid, fpca_mod2$efunctions[, 4], type="l", xlab="bin", ylab="PC4",
     ylim = c(-0.14, 0.23))
```

- Individual covariance matrix

Wow! This covariacne matrix changed a lot from when using all of the PCs! It still looks cyclic, but now off-diagnoal region seems much more stable. Fluctuations mostly centered around diagnoal line.

```{r}
# covariance
heatmap(fpca_mod$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
        Rowv = NA, Colv = NA, main = "glmer")

heatmap(fpca_mod2$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
        Rowv = NA, Colv = NA, main = "glmmTMB")

## Dynamic prediction

Still from last time, we're gonna focus on Bayes methods now!

We are also going to extend the observed track, so that there are some variation in the observed data (instead of all zeros). I am going to start with 9am (540), 1pm (780),  5pm (1020)

Extending the track seems to solve the problem with failed Laplace Approximation. Not sure. Should try on the full sample.

```{r, warning=FALSE, message=FALSE}
df_pred <- df_est_latent %>% filter(id %in% rand_id)
df_pred[, 'pred_t540'] <- df_pred[, 'pred_t780'] <- df_pred[, 'pred_t1020'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
```

```{r, results='hide'}
# unique id
# prediction for a single subject
source(here("Code/OutsampBayes.R"))

# PRILS
for(i in rand_id){
  df_i <- df %>% filter(id==i)

  # prediction 
  ## up to 360
  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=540), kpc=K)
  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t1$score_out
  df_pred[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
  ## up to 720
  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=780), kpc=K)
  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t2$score_out
  df_pred[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
  ## up to 1080
  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1020), kpc = K)
  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t3$score_out
  df_pred[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3

}

df_pred$pred_t540[df_pred$bin<=540] <- NA
df_pred$pred_t780[df_pred$bin<=780] <- NA
df_pred$pred_t1020[df_pred$bin<=1020] <- NA
```

```{r, results='hide'}
# glmmTMB
df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
df_pred2[, 'pred_t540'] <- df_pred2[, 'pred_t780'] <- df_pred2[, 'pred_t1020'] <- NA

for(i in rand_id){
  df_i <- df %>% filter(id==i)

  # prediction 
  ## up to 360
  pred_t1 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=540), kpc=K)
  eta_pred_t1 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t1$score_out
  df_pred2[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
  ## up to 720
  pred_t2 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=780), kpc=K)
  eta_pred_t2 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t2$score_out
  df_pred2[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
  ## up to 1080
  pred_t3 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=1020), kpc = K)
  eta_pred_t3 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t3$score_out
  df_pred2[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3

}

df_pred2$pred_t540[df_pred$bin<=540] <- NA
df_pred2$pred_t780[df_pred$bin<=780] <- NA
df_pred2$pred_t1020[df_pred$bin<=1020] <- NA
```

```{r, warning=FALSE, fig.width=16, fig.height=8}
ggarrange(
df_pred %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
  facet_wrap(~id,scales = "free")+labs(title = "glmer"),

df_pred2 %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
  facet_wrap(~id,scales = "free")+labs(title = "glmmTMB"),
nrow = 1, common.legend = T)
```

# NHANES data application output

## fGFPCA

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
- Used PIRLS for local GLMMs step

```{r}
load(here("Data/ApplOutput_fGFPCA.RData"))
```

```{r}
# linear interpolation
# remove subjects with numeric problem
df_test_full <- df %>% 
  filter(id %in% unique(df_test$id)) %>%
  filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
df_test_full$pred_t540_interp <- df_test_full$pred_t780_interp <- df_test_full$pred_t1020_interp <- NA

# interpolation for each subject
t1 <- Sys.time()
for(i in unique(df_test_full$id)){
  # interpolation t540
  interp_i_t540 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t540"],
                     xout = 541:J, rule = 2)
  df_test_full$pred_t540_interp[df_test_full$id==i & df_test_full$sind>540] <- interp_i_t540$y

  # interpolation t780
  interp_i_t780 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t780"],
                     xout = 781:J, rule = 2)
  df_test_full$pred_t780_interp[df_test_full$id==i & df_test_full$sind>780] <- interp_i_t780$y

  # interpolation t1020
  interp_i_t1020 <- approx(x = df_test[df_test$id==i, "bin"], 
                     y = df_test[df_test$id==i, "pred_t1020"],
                     xout = 1021:J, rule = 2)
  df_test_full$pred_t1020_interp[df_test_full$id==i & df_test_full$sind>1020] <- interp_i_t1020$y
}
t2 <- Sys.time()
t_interp <- t2-t1

- `r round(t_local_glmm, 2)` minutes spent on fitting local GLMMs, `r round(t_fpca, 2)` seconds on FPCA model, `r round(t_pred, 2)` minutes on out-of-sample prediction, `r round(t_interp, 2)` minutes on interpolation to the original grid

- Let's look at four subjects as an example:

```{r message=FALSE, warning=FALSE}
rand_id2 <- sample(unique(df_test$id), 4)
# without interpolation, assume constant latent function value in each bin
df_test_full%>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
         title = "Constant latent function within bin")
```

### A closer look at interpolation

- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints. Here I wanna compare if this choice would make any different. Also, there are a lot of other methods to fill in the gaps, such as projection. 
- For linear interpolation, I set the edges to be the closest data extreme.
- The interpolation itself spent `r round(t_interp, 2)` minutes.

```{r, warning=FALSE}
# with interpolation
df_test_full%>% 
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540_interp, pred_t780_interp, pred_t1020_interp), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540_interp, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780_interp, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020_interp, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
         title = "Interpolated latent function")

@@

- Three subjects had failed approximation, all happened with the shortest observed track
- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width).

@@

```{r}
# with interpolation
# up to 9am
auc_t540_interp <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540_interp, x$Y), measure = "auc"))
auc_t540_interp <- lapply(auc_t540_interp, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1pm
auc_t780_interp <- lst_auc[3:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t780_interp, x$Y), measure = "auc"))
auc_t780_interp <- lapply(auc_t780_interp, function(x){x@y.values[[1]]}) %>% unlist()

# up to 5pm
auc_t1020_interp <- lst_auc[4] %>% 
  lapply(function(x)performance(prediction(x$pred_t1020_interp, x$Y), measure = "auc"))
auc_t1020_interp <- lapply(auc_t1020_interp, function(x){x@y.values[[1]]}) %>% unlist()
```

@@

- Below is the model fit on the trainign set

@@

idx1 <- idx2 <- idx3 <- c(1, 3, 3, 3)
names(idx1) <- c("Total time", as.character(round(t_total, 2)))
names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))

>>>>

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
theme_set(theme_minimal())

@@

df %>% 
  filter(SEQN %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Z), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
```
## fGFPCA

```{r}
# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))
```

### Bin Data

- Bin every 10 consecutive observations

@@

df %>% 
  filter(SEQN %in% rand_id) %>%
  group_by(SEQN, bin) %>%
  summarise(num = sum(Z)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")

df <- df %>% rename(id = SEQN, Y=Z)
```
### Data split

```{r}
id_vec <- unique(df$id)
train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
```

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction

### Local GLMM

- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
- Fit on the training set

```{r, cache=TRUE}
# fit model on the training set
df_train <- df %>% filter(id %in% train_id)
train_bin_lst <- split(df_train, f = df_train$bin)

# local GLMM and estimate latent function
# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
t_local_glmm <- t2-t1 
```

```{r}
df_est_latent <- bind_rows(df_est_latent) %>% 
  select(-sind, -Y) %>% distinct(.) 
# example estimated latent function
df %>% 
  filter(id %in% sample(train_id, size = 4)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r, message=FALSE}

df %>% 
  filter(id %in% sample(train_id, size = 4)) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  facet_wrap(~id)+
  labs(x="Time", y = "Estimated latent function (original scale")
```

```{r}
mat_est_unique <- matrix(df_est_latent$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) # row index subject, column binned time
#dim(mat_est_unique)

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model

# dim(fpca_mod$efunctions) # 27 eigenfunctions total
# fpca_mod$evalues
K <- 4
```

```{r}
plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
```

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,2))

plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

```{r}
# plot correlation matrix
sum(fpca_mod$VarMats[[1]]!=fpca_mod$VarMats[[2]]) 
heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
```

### Debias?

### Out-of-sample prediction

follow up on the last report, I'll add some details here.

Let's set:

- $s$: bin index
- $t_{m_s}$: midpoint time of bin s
- $\mathcal{T}_s$: all observation points in bin s
- $Y_i^s)$ all observations in bin s from subject i

Let's assume we have a new subject $u$ with maximum observation time $T_u$. Then the log-likelihood of this new subject would be:

$$
l_u=\sum_{t_{m_s}<T_u}log(h(Y_u^s))+\hat{\eta}_u(t_{m_s})T(Y_u^s)-log(A[\hat{\eta}_u(t_{m_s})])
$$

where $\hat{\eta}_u(t_{m_s}) = \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})$.

From FPCA model (or the debias step to be added), we obtain $\hat{\mu}_0$, $\hat{\phi}_k$, as well as the variance estimates of $\xi_k$: $\hat{\lambda}_k$, and variance of residual process $\hat{\sigma}^2$.

With all these estimates, we wanna use Laplace approximation to find the $\xi_{uk}$ that maximizes $l_u$.

I am using a package called **LaplacesDemon**, following this example https://cran.r-project.org/web/packages/LaplacesDemon/vignettes/LaplacesDemonTutorial.pdf

```{r, class.source = "fold-show"}
library(LaplacesDemon)
```

Now I am going to use one subject as an example to go over the approximation procedure:

First we write out the model:

- Prior distribution: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$
- Posterior distribution:

For a single bin
$$
\begin{aligned}
l(\mathbf{Y_u}|\mathbf{\xi}_u) &= \sum_{t_{m_s}<T_u}l(Y_u^s|\mathbf{\xi}_u) \\

Y_u^s|\mathbf{\xi}_u & \sim Binomial(n_s, p_s) \\

g^{-1}(p_s) = \hat{\eta}(t_{m_s}) &= \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})
\end{aligned}
$$
And $n_s$ is the number of observations in bin s.

Bayes theroem:

$$
l(\mathbf{\xi}_u|\mathbf{Y}_u) \propto l(\mathbf{Y}_u|\mathbf{\xi}_u)+l(\mathbf{\xi}_u)
$$

```{r, class.source = "fold-show"}
## model
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

  # log-prior
  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)

  # log-posterior likelihood
  eta <- Data$f0+Data$X %*% xi
  p <- exp(eta)/(1+exp(eta))
  LL <- sum(dbinom(x=Data$y, size = Data$n, prob=p, log = TRUE)) # log likelihood of Y|xi
  LP <- LL+sum(xi.prior) # unnormalized joint log likelihood of (Y, xi)

  # output
  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LP,
                   yhat=Data$y, parm=parm)
  return(Modelout)
}
```

- Then we feed in data for one observation

```{r, class.source = "fold-show"}
rand_id <- sample(test_id, 1)
df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am

# put data into correct format
ns <- as.vector(table(df_new$bin)) # number of observations
hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
nf <- ns-hs # number of failure
max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)

# extract estimates from FPCA model to be used in prior distribution
tao <- diag(fpca_mod$evalues[1:K]) # variance of xi

# extract estimates from FPCA model to be used in posterior distribution
f0 <- fpca_mod$mu[1:max_bin]
N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
n <- df_new2$ns # number of experiements at each bin
y <- df_new2$hs # outcome, number of 1
X <- fpca_mod$efunctions[1:max_bin, 1:K] # eigen/PC functions
J <- K # number of parameters/scores
mon.names <- "LP"

# parameter names
name_lst <- as.list(rep(0, J))
names(name_lst) <- paste("xi", 1:J, sep = "")
parm.names <- as.parm.names(name_lst) # names of parameters to estimate
pos.xi <- grep("xi", parm.names)

# data
MyData <- list(J=J, X=X, mon.names=mon.names,
                 parm.names=parm.names, pos.xi=pos.xi, y=y, n=n, tao=tao, f0=f0)
```

```{r, class.source = "fold-show"}
# fit laplace approximation
# initial values generated by GIV function in the same package
Fit <- LaplaceApproximation(Model, Data=MyData)
```

- Summarizing output

```{r, class.source = "fold-show", fig.height=8, fig.width=8}
# summarizing output
print(Fit)
plot(Fit, MyData)
```

```{r}
# prediction
eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%Fit$Summary1[, "Mode"]

data.frame(bin = mid, eta_pred = eta_pred_out) %>% 
  right_join(df %>% filter(id==rand_id) %>%  select(Y, sind, bin), by = "bin") %>%
  ggplot()+
  geom_point(aes(x=sind, y = Y))+
  geom_line(aes(x=sind, y = exp(eta_pred)/1+exp(eta_pred)))+
  labs(title = rand_id)

@@

## Full data application output

### fGFPCA

- Still need to project eigenfunctions to the full grid

```{r}
# results generated from DataApplNHANES.R
load(here("Data/ApplOutput_fGFPCA.RData"))
```

```{r message=FALSE, warning=FALSE}
df_test_full <- df %>% 
  filter(id %in% test_id) %>%
  filter(!id %in% skip_id) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))

rand_id2 <- sample(unique(df_test$id), 4)
# without interpolation, assume constant latent function value in each bin
df_test_full%>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)",
         title = "Constant latent function within bin")
```

@@

- Three subjects had failed approximation, all happened with the shortest observed track
- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width). 
- Per last meeting, adding some details on laplace approximation

@@

@@

- Below is the model fit on the training set

@@

idx1 <- idx2 <- idx3 <- c(1, 3, 3)
names(idx1) <- c("Total time", as.character(round(t_total, 2)))
names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))
